SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2019-03-06 scipy-1.0/paper.tex
removed commented out figure

Python, Cython, Fortran, C and C++ are the programming languages used to
implement scientific algorithms in the SciPy library. An analysis of our code
base using the \texttt{linguist} library\cite{linguistref} provides a 
detailed breakdown as \% composition by programming language in 
SciPy (Figure~\ref{fig:linguist}).

\begin{figure}[H]
    \centering
		\includegraphics[width=\textwidth]{static/composition}
    \caption{The breakdown of programming languages used in the
             SciPy library determined using the linguist library.
    	 Small ($<0.5 \%$) amounts of TeX, Matlab, Shell,
    	 and Makefile are excluded for clarity and mostly
    	 provide supporting roles in tests, building, and
    	 documentation.}
    \label{fig:linguist}
\end{figure}

For implementing new functionality, we have a clear order of language
preference.  First Python, if performance is not an issue. If it is, then in
order of preference: Cython, C, C++, Fortran. The main motivation for this is
maintainability: Cython has the highest abstraction level and most Python
developers will understand it. C is also widely known, and easier to deal with
for the core development team than C++ and especially Fortran.

While it is not surprising that Python is heavily used in SciPy, the usage
distribution of the other languages warrants some discussion. Fortran is an
extremely well-established scientific programming language, both for historical
reasons and because of its continued excellent
performance\cite{Koelbel:1993:HPF:562354}. We wrap the FFTPACK Fortran library
for performing Fourier transforms\cite{SWARZTRAUBER198445, SWARZTRAUBER198251}
since this library has been a standard in the field for 33 years and has a
license that is compatible with our own. Likewise, we wrap the Fortran source
for ODEPACK\cite{citeulike:2644528} as it has been  trusted for the initial
value problem for ordinary differential equation systems for 30 years. For
similar reasons we also wrap the Fortran libraries
QUADPACK\cite{1983qspa.book.....P} (for numerical integration of
one-dimensional functions), FITPACK\cite{Dierckx:1993:CSF:151103}
(curve-fitting / interpolation), ODRPACK\cite{ODRPACK_Boggs} (orthogonal
distance regression), MINPACK\cite{osti_6997568} (minimization of linear and
nonlinear equations), ARPACK\cite{leh:sor:yan96} (solving large scale
eigenvalue problems), ALGORITHM 644\cite{Amos:1986:APP:7921.214331} (handling
Bessel Functions), and CDFLIB\cite{CDFLIB_site} (evaluation of cumulative
density functions).

Rounding out the top three languages in SciPy is C, which is also extremely
well-established over several decades\cite{Kernighan:1988:CPL:576122} of
scientific computing. Alongside its objected-oriented relative C++, C can be
leveraged in Python after being exposed using the Cython language. Cython has
been described as a creole language that mixes the best parts of Python and
lower-level C / C++ paradigms\cite{behnel2011cython}. We thus often use Cython
as a glue between well-established low-level scientific computing libraries and
the Python interface offered by SciPy. We also use Cython to enable performance
enhancements in Python code, especially for cases where heavily used inner
loops benefit from a compiled code with static typing. Some of the C libraries
that we wrap in SciPy include trlib\cite{doi:10.1080/10556788.2018.1449842}
(iterative solving of the trust region problem),
SuperLU\cite{li05,superlu_ug99} (solution of large, sparse, nonsymmetric
systems of linear equations), Qhull\cite{Barber:1996:QAC:235815.235821}
(computational geometry), and Cephes\cite{cephes_netlib} (mathematics
algorithms).

Therefore, the relative abundance of different programming languages in the
SciPy library results from a combination of the usage of powerful performance
enhancing languages that interact well with Python (i.e., Cython), and the
usage of languages (and their libraries) that have proven reliable and
performant over many decades. The position that SciPy occupies near the
foundation of the scientific Python ecosystem is such that adoption of new
languages or major dependencies is generally unlikely---our choices are strongly
driven by long-term stability. GPU acceleration, new transpiling libraries, and
the latest JIT compilation approaches (i.e.,
Numba\cite{Lam:2015:NLP:2833157.2833162}) are very powerful, but currently fall
outside the remit of the main SciPy library. That said, we have recently
increased our efforts to support compatibility with some of these options, and
having our full test suite pass with the PyPy JIT
compiler\cite{Bolz:2009:TMP:1565824.1565827} is now a requirement in our
development workflow.

@@

\subsubsection*{\texttt{cKDTree}}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{static/asv_time_query_ckdtree}
\caption{Airspeed velocity benchmarks for \texttt{scipy.spatial.cKDTree.query()}
over a roughly nine year commit history time frame. The results are based on
Python 2.7 performance on the master branch of the project using NumPy 1.8.2
and Cython versions 0.27.3, 0.21.1, and 0.18 (for improved backward
compatibility). Only the L2 (Euclidean) norm is shown here, and to improve
backward compatibility and sampling of the benchmarks there was no application
of toroidal topology to the \texttt{KDTree} (boxsize argument was ignored).}
\label{fig:asvbench}
\end{figure}

@@

The SciPy test suite is orchestrated by a continuous integration matrix that
includes POSIX and Windows (32/64-bit) platforms managed by Travis CI and
AppVeyor, respectively. Our tests cover Python versions 2.7, 3.4, 3.5, 3.6, and
include code linting with \texttt{pyflakes} and \texttt{pycodestyle}. There are more than $13,000$
unit tests in the test suite, which is written for usage with the \texttt{pytest}
framework, and with 87 \% and 45 \% line coverages for Python and compiled
code, respectively at the SciPy 1.0 release point (Figure~\ref{fig:coverage}). Documentation for the code is automatically built and hosted by
the CircleCI service to facilitate evaluation of documentation changes /
integrity.  Our full test suite also passes with PyPy3, a just-in-time compiled
version of the Python language.

\begin{figure}[H]
\centering
\begin{tikzpicture}[]
\pgfplotstableread[column/ver/.style=string type]{
ver totpyline covpyline uncovpyline totcompline covcompline uncovcompline
v0.12.1 33844 25721 8123 273634  142290 131344
v0.13.3 36638 28944 7694 338926  172852 166074
v0.14.1 37301 30587 6714 277643  147151 130492
v0.15.1 38339 30288 8051 304898  158547 146351
v0.16.1 40094 32075 8019 322398  167647 154751
v0.17.1 42566 34478 8088 340903  170452 170451
v0.18.1 44711 36216 8495 353417  173174 180243
v0.19.1 43823 36373 7450 436900  200974 235926
v1.0.0 106878 92984 13894 462574 208158 254416
}\covtable
    \begin{axis}[
    width=\textwidth,
    height=7cm,
    enlargelimits=false,
    ymin=0,
    ymax=6e5,
    stack plots=y,%   
    area style,
    xtick=data,
    ytick={2e5,4e5,6e5},
    xticklabels from table={\covtable}{ver},
    scaled y ticks=false,
    y tick label style={/pgf/number format/.cd, fixed, 1000 sep={}},
    legend entries={Compiled (covered),Compiled (uncovered),Python (covered),Python (uncovered)},
    legend style={draw=none,fill=none, cells={align=left},/tikz/every even column/.append style={column sep=3mm}},
    legend columns=-1,
    legend image code/.code={\fill[#1] (0cm,-0.1cm) rectangle (0.2cm,0.1cm);},
    ]
  \addplot+[fill=blue!50] table[x expr=\coordindex, y=covcompline] {\covtable} \closedcycle;
  \addplot+[fill=blue!10] table[x expr=\coordindex, y=uncovcompline] {\covtable} \closedcycle;
  \addplot+[fill=red!70] table[x expr=\coordindex, y=covpyline] {\covtable} \closedcycle;
  \addplot+[fill=red!30] table[x expr=\coordindex, y=uncovpyline] {\covtable} \closedcycle;
\end{axis}
\end{tikzpicture}
\caption{
Python (red) and compiled (blue) code volume in SciPy over time.
Deep-shaded area represents lines of code covered by units tests;
light-shaded area represents lines not covered. With the exception
of the removal of $\approx 61,000$ lines of compiled code for SciPy
v0.14, the volume of both compiled and Python
code has increased between releases, as has the number of lines
covered by unit tests. SciPy v1.0.0 was composed of $462,574$ lines of
compiled code with test coverage near $45 \%$ and $106,878$ lines of 
Python code with test coverage near $87 \%$.\\
The reconstitution of historical
build and test environments was performed using a docker-based approach
\cite{scipy-cov}. Analysis of lines of source code covered is performed 
automatically in our continuous integration suite using the \texttt{pytest-cov}
and \texttt{gcov} libraries. Some decreases in compiled code coverage may reflect
inclusion of additional vendored code.}
\label{fig:coverage}
\end{figure}

@@

SciPy adopted an official Governance Document on August 3,
2017\cite{SciPyProjectGovernance}.  We have a Benevolent Dictator for Life
(BDFL), Pauli Virtanen, leading the project, following a detailed discussion of
the governance model with the community on the mailing list.  In practice, the
overruling authority of the BDFL is anticipated to be used only when the
Steering Council cannot agree on a matter.  The BDFL is expected to consider
any strong indications that they should step down; although the BDFL may
appoint a successor, it is generally expected that the Steering Council will be
consulted on the matter.

@@

\begin{table}[H]
\caption{Summary of SciPy Roadmap items following 1.0 release}
\begin{tabular}{ c|c }
    SciPy submodule & Summary of Change \\
    \hline
    \texttt{optimize} & a few more high quality global optimizers\\
    \texttt{fftpack} & reduce overlap with NumPy equivalent \\
    \texttt{linalg} & reduce overlap with NumPy equivalent \\
    \texttt{interpolate} & new spline fitting and arithmetic routines \\
    \texttt{interpolate} & new transparent tensor-product splines\\
    \texttt{interpolate} & new Non-Uniform Rational B-Splines\\
    \texttt{interpolate} & mesh refinement \& coarsening of B-splines and tensor products\\
    \texttt{signal} & migrate spline functionality to \texttt{interpolate}\\
    \texttt{signal} & Second Order Sections update to match capabilities in other routines\\
    \texttt{linalg} & support a more recent version of LAPACK\\
    \texttt{ndimage} & clarify usage of the ``data point'' coordinate
    model, and add additional wrapping modes\\
    \texttt{sparse} & incorporate sparse arrays from Sparse package\cite{abbasi2018sparse} \\
    \texttt{sparse.linalg} & add PROPACK wrappers for faster SVD\\
    \texttt{spatial} & add support for (quaternion) rotation matrices\\
    \texttt{special} & precision improvements for hypergeometric, parabolic cylinder, and spheroidal wave functions\\
\end{tabular}\label{table:roadmap}
\end{table}

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{supporting_info/knn_complexity_confirm}
\caption{Assessment of \texttt{cKDTree} nearest neighbor query time complexity as a function of the number of input points (N) along with similar assessment for related algorithms in scikit-learn. The calculations are based on code previously reported\cite{knn-jake}, but updated to use SciPy 1.0.0, Python 3.6, NumPy 1.13.1, and Cython 0.27.3.}
\label{fig:knn-complexity}
\end{figure}

@@

\section{Polynomial Interpolators: Additional Details}
FITPACK provides a collection of routines for calculating
smoothing splines and automatic knot selection. One issue 
with FITPACK and its Python wrappers in SciPy, \code{UnivariateSpline}
and the \code{splrep}/\code{splev} combination, is the monolithic design:
specific algorithms for construction of interpolating and 
smoothing splines and knot selection are tightly coupled 
at the Fortran level. The library offers a very
limited degree of user control over the details 
of the spline fitting it performs. 
%
Routines for manipulating piecewise polynomials and b-splines were either
not available or lacking in ease of use and/or performance characteristics.

An alternative set of spline interpolation routines was implemented in the
(now deprecated) \code{splmake}/\code{spleval} combination, and used as a
backend of the \code{interp1d} routine. To further complicate matters,
the spline format used by \code{splmake}/\code{spleval} was incompatible 
with \code{splrep}/\code{splev}.

Using \code{UnivariateSpline} in new code is not recommended. Given data points,
the helper function \code{make\_interp\_spline} constructs a \code{BSpline}
instance, representing a spline which interpolates the data. Interpolated values
at intermediate points can be computed either via its \code{\_\_call\_\_} method,
or by calling \code{splev(xnew, BSpline\_object)} --- the \code{splev} routine
accepts both \code{tck} tuples and \code{Bspline} instances.

Fitting \emph{smoothing} splines to data can currently only be done with 
FITPACK. Building alternative algorithms is certainly possible,
but is still to be implemented.

Interpolation with known or estimated derivatives, i.e., cubic or
higher-order Hermite splines, can be constructed by either
subclassing \code{PPoly} or \code{BPoly}, or using the factory
method \code{BPoly.from\_derivatives}.

The spline representation used by \code{BSpline} is directly
compatible with FITPACK. We recommend that users work with
\code{BSpline} objects instead of directly manipulating \code{tck}
tuples of knots, coefficients and the spline degree.

@@

    can deal with constraints on the variables. We distinguish among \textit{bound constraints} (i.e. $x^l \le x \le x^u$),

>>>>

Figure~\ref{fig:linguist} provides a breakdown of the programming languages used
in SciPy. While it is not surprising that Python is predominant, the usage distribution 
of the other languages warrants discussion.

\begin{figure}[H]
    \centering

    \begin{tikzpicture}[font=\sffamily]
    \begin{axis}[height=6cm, width=10cm,
    ybar, ymax=60, ymin=0,bar width=1cm,
    ylabel={\% of Scipy code base},
    ytick={0,20,...,60},
    axis lines*=left,
    axis y line shift=0pt,
    enlarge x limits=0.15,
    symbolic x coords={py,f,c,cy,cpp},
    xticklabels={Python,FORTRAN,C,Cython,C++},
    ticklabel style={align=center,font=\sffamily\small,
    /pgf/number format/assume math mode=true},
    xtick=data,
    nodes near coords,
    nodes near coords style={/pgf/number format/assume math mode=true}
    ]
    \addplot coordinates {(py,49.5) (f,25.6) (c,19.5) (cy, 3.0) (cpp,2.3)};
    \end{axis}
    \end{tikzpicture}
    \caption{The breakdown of programming languages used in the
             SciPy library determined using the \texttt{linguist} library.
    	 Small ($<0.5 \%$) amounts of TeX, Matlab, Shell,
    	 and Makefile are excluded for clarity and mostly
    	 provide supporting roles in tests, building, and
    	 documentation.}
    \label{fig:linguist}
\end{figure}

Fortran is an
extremely well-established scientific programming language, both for historical
reasons and because of its continued excellent
performance\cite{Koelbel:1993:HPF:562354}. We wrap the FFTPACK Fortran library
for performing Fourier transforms\cite{SWARZTRAUBER198445, SWARZTRAUBER198251}
since this library has been a standard in the field for 33 years and has a
license that is compatible with our own. Likewise, we wrap the Fortran source
for ODEPACK\cite{citeulike:2644528} as it has been  trusted for the initial
value problem for ordinary differential equation systems for 30 years. For
similar reasons we also wrap the Fortran libraries
QUADPACK\cite{1983qspa.book.....P} (for numerical integration of
one-dimensional functions), FITPACK\cite{Dierckx:1993:CSF:151103}
(curve-fitting / interpolation), ODRPACK\cite{ODRPACK_Boggs} (orthogonal
distance regression), MINPACK\cite{osti_6997568} (minimization of linear and
nonlinear equations), ARPACK\cite{leh:sor:yan96} (solving large scale
eigenvalue problems), ALGORITHM 644\cite{Amos:1986:APP:7921.214331} (handling
Bessel Functions), and CDFLIB\cite{CDFLIB_site} (evaluation of cumulative
density functions).

Rounding out the top three languages in SciPy is C, which is also extremely
well-established over several decades\cite{Kernighan:1988:CPL:576122} of
scientific computing. Some of the C libraries
that we wrap in SciPy include trlib\cite{doi:10.1080/10556788.2018.1449842}
(iterative solving of the trust region problem),
SuperLU\cite{li05,superlu_ug99} (solution of large, sparse, nonsymmetric
systems of linear equations), Qhull\cite{Barber:1996:QAC:235815.235821}
(computational geometry), and Cephes\cite{cephes_netlib} (mathematics
algorithms).

Cython has been described as a creole language that mixes the best parts of Python and
lower-level C / C++ paradigms\cite{behnel2011cython}. We often use Cython
as a glue between well-established low-level scientific computing libraries written
in C / C++ and the Python interface offered by SciPy. We also use Cython to enable performance
enhancements in Python code, especially for cases where heavily used inner
loops benefit from a compiled code with static typing.

For implementing new functionality, Python is the still the language of choice. If Python 
performance is an issue, then we prefer the use of Cython followed by C, C++, or Fortran (in that
order). The main motivation for this is maintainability: Cython has the 
highest abstraction level and most Python developers will understand it. C is 
also widely known, and easier for the current core development team to manage
than C++ and especially Fortran.

The distribution of secondary programming languages in SciPy is a compromise between
a powerful, performance-
enhancing language that interacts well with Python (i.e. Cython) and the
usage of languages (and their libraries) that have proven reliable and
performant over many decades. The position that SciPy occupies near the
foundation of the scientific Python ecosystem is such that adoption of new
languages or major dependencies is generally unlikely---our choices are strongly
driven by long-term stability. GPU acceleration, new transpiling libraries, and
the latest JIT compilation approaches (i.e.,
Numba\cite{Lam:2015:NLP:2833157.2833162}) are very powerful, but currently fall
outside the remit of the main SciPy library. That said, we have recently
increased our efforts to support compatibility with some of these options, and
having our full test suite pass with the PyPy JIT
compiler\cite{Bolz:2009:TMP:1565824.1565827} is now a requirement in our
development workflow.

@@

\subsubsection*{cKDTree}

@@

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{static/asv_time_query_ckdtree}
\caption{Airspeed velocity benchmarks for scipy.spatial.cKDTree.query()
over a roughly nine year commit history time frame. The results are based on
Python 2.7 performance on the master branch of the project using NumPy 1.8.2
and Cython versions 0.27.3, 0.21.1, and 0.18 (for improved backward
compatibility). Only the L2 (Euclidean) norm is shown here, and to improve
backward compatibility and sampling of the benchmarks there was no application
of toroidal topology to the KDTree (boxsize argument was ignored).}
\label{fig:asvbench}
\end{figure}

@@

The SciPy test suite is orchestrated by a continuous integration matrix that
includes posix and Windows (32/64-bit) platforms managed by Travis CI and
appveyor, respectively. Our tests cover Python versions 2.7, 3.4, 3.5, 3.6, and
include code linting with pyflakes and pycodestyle. There are more than $13000$
unit tests in the test suite, which is written for usage with the pytest
runner, and with a 76.5 \% coverage of approximately $204000$ lines of code
holding steady for the last six months (Figure~\ref{fig:coverage}).  Some of
the historical components of the code base would still benefit from increased
test coverage. Documentation for the code is automatically built and hosted by
the CircleCI service to facilitate evaluation of documentation changes /
integrity.  Our full test suite also passes with PyPy3, a just-in-time compiled
version of the Python language.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{static/coverage-chart}
\caption{\% coverage of the SciPy code base by unit tests over the six months
preceding June 28, 2018 as reported by the \texttt{codecov} service. The
analysis of lines of source code covered is performed automatically in our
continuous integration suite using the \texttt{pytest-cov} and \texttt{gcov}
libraries.}
\label{fig:coverage}
\end{figure}

@@

SciPy adopted an official Governance Document on August 3,
2017\cite{SciPyProjectGovernance}.  We have a Benevolent Dictator for Life
(BDFL), Pauli Virtanen, leading the project, following a detailed discussion of
the governance model with the community on the mailing list.  In practice, the
overruling authority of the BDFL is anticipated to be used only when the
Steering Council cannot agree on a matter.  The BDFL is expected to consider
any strong indications that they should step down; although the BDFL may
appoint a successor it is generally expected that the Steering Council will be
consulted on the matter.

@@

\begin{table}[H]
\caption{Summary of SciPy Roadmap items following 1.0 release}
\begin{tabular}{ c|c }
    SciPy submodule & Summary of Change \\
    \hline
    \texttt{optimize} & a few more high quality global optimizers\\
    \texttt{fftpack} & reduce overlap with NumPy equivalent \\
    \texttt{linalg} & reduce overlap with NumPy equivalent \\
    \texttt{interpolate} & new spline fitting and arithmetic routines \\
    \texttt{interpolate} & new transparent tensor-product splines\\
    \texttt{interpolate} & new Non-Uniform Rational B-Splines\\
    \texttt{interpolate} & mesh refinement \& coarsening of B-splines and tensor products\\
    \texttt{signal} & migrate spline functionality to \texttt{interpolate}\\
    \texttt{signal} & Second Order Sections update to match capabilities in other routines\\
    \texttt{linalg} & support a more recent version of LAPACK\\
    \texttt{ndimage} & migrate to data point model from pixel model\\
    \texttt{sparse} & incorporate sparse arrays from Sparse package\cite{abbasi2018sparse} \\
    \texttt{sparse.linalg} & add PROPACK wrappers for faster SVD\\
    \texttt{spatial} & add support for (quaternion) rotation matrices\\
    \texttt{special} & precision improvements for hypergeometric, parabolic cylinder, and spheroidal wave functions\\
\end{tabular}\label{table:roadmap}
\end{table}

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{supporting_info/knn_complexity_confirm}
\caption{Assessment of cKDTree nearest neighbor query time complexity as a function of the number of input points (N) along with similar assessment for related algorithms in scikit-learn. The calculations are based on code previously reported\cite{knn-jake}, but updated to use SciPy 1.0.0, Python 3.6, NumPy 1.13.1, and Cython 0.27.3.}
\label{fig:knn-complexity}
\end{figure}

@@

@@

    can deal with constraints on the variables. We distinguish between: \textit{bound constraints} (i.e. $x^l \le x \le x^u$),
